Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-10-16 manuscript/climate_sensitivity_SI.Rmd
Moved SI to another .rmd file

---
title: "Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [book.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
), 'packages.bib')
```
\raggedright

### Title page

text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	## (limit 200) | 	No. of Tables | 	5  
Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
Materials and Methods |	currently ~1442	||	 
Results |	currently ~1437	 ||
Discussion |	currently ~1881 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | 	~75 unwritten	||

\newpage

### Summary (**needs work**)
- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important. Yet it remains unclear how tree size and species' traits interactively shape tree growth responses across droughts.

- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60 year period from 1950 and 2009.

- Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration). However, there was substantial variation in the best predictor variables across given droughts.

- We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.

*The Summary for research papers, which must be usable as a stand- alone document, must not exceed 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion. It should not contain citations of other papers.*

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these. Very short phrases and scientific names with their common equivalents (e.g. Nicotiana tabacum (tobacco)) are acceptable.*

\newpage
### Introduction
Forests globally play a critical role in climate regulation ([Bonan 2008](https://science.sciencemag.org/content/320/5882/1444)), yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change ([Friedlingstein et al. 2006](https://journals.ametsoc.org/doi/full/10.1175/JCLI3800.1)). An important aspect of this uncertainty lies in responses to drought (**REF**). In many forested regions around the world, the risk of severe drought is increasing ([Trenbert et al. 2014](https://www.nature.com/articles/nclimate2067)), even in conjunction with increasing precipitation ([IPCC 2014](https://www.cambridge.org/core/books/climate-change-2014-impacts-adaptation-and-vulnerability-part-b-regional-aspects/036A899BD52861D61B0D519C5F2B9334)). Global change-type drought has been affecting forests worldwide ([Allen et al. 2015](https://doi.org/10.1016/j.foreco.2009.09.001)), and it is expected that future climate change-driven droughts will severely impact forests around the world ([Allen et al. 2015](https://doi.org/10.1016/j.foreco.2009.09.001); **REFS**). Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [Stovall et al. 2019]( https://doi.org/10.1038/s41467-019-12380-6)), resulting in disproportionate impacts on forest C storage ([Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear. Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al. 2019, Hacket-Pain et al. 2016 (DOI 10.1007/s10342-016-0982-7) **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (*Pfeifer, E. M., Hicke, J. A. & Meddens, A. J. H. Observations and modeling of aboveground tree carbon stocks and fluxes following a bark beetle outbreak in the western United States. Glob. Change Biol. 17, 339–350 (2011).*)--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (* McDowell, N. G., Bond, B. J., Hill, L., Ryan, M. G. & Whitehead, D. in Size and age related changes in tree structure and function (eds Meinzer, F.C. & Niinemets, U.) 255–286 (Springer Publishing, 2011); McDowell, N. G. & Allen, C. D. Darcy’s law predicts widespread forest mortality under climate warming. Nat. Clim. Change 5, 669–672 (2015); Ryan, M. G., Phillips, N. & Bond, B. J. The hydraulic limitation hypothesis revisited. Plant Cell Environ. 29, 367–381 (2006).*), and this may become a greater liability during drought ([Zhang et al. 2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)). Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al. 2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al. 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)). Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses. Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019](Elliot et al. 2015, Friedrichs et al. 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**). However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al. 2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al. 2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)). There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al. 2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested. (**But see D'Orangeville et al. 2018--except that it may not be a well-constructed test for traits**)

**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand ([D'Orangeville et al. 2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14096)). However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts. (**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011];  D'Orangeville et al. 2018 **REFS**), these don't test for differential trait effects across droughts (D'Orangeville et al. 2018) and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment. The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [Allen et al. 2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**). Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited. While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**). There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; Stovall et al. 2019). [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
(**PELTIER**)

Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses. 
First, we focus on the role of tree size and its interaction with microenvironement. 
We confirm that, consistent with most forests globally, larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*). 
We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*). 
Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height. We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*). 
We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).

```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

\clearpage

### Materials and Methods
*Study site*

Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al. 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

*Data collection and preparation*

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\le$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2018 (**confirm**), are available through the [ForestGEO data portal](www.forestgeo.si.edu).

```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
  column_spec(4, width = "7cm") %>%
  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  kable_styling()
```

We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019]. Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  *Ryan submitted the data to ITRDB but I don't think its posted yet. We should also cite GitHub/Zenodo here. I'll come back to that.*

For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
Prior DBH was estimated using the following equation:

$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$

Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#) and then used these to estimate bark thickness based on DBH.

Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013]. Species-specific height allometries were developed (Table S#). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.

Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed. As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years. While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig. 2) and height change since *1959* indicated that change was likely slow. Specifically, [**Ian, please provide details--e.g., average rate of height growth**]

Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [** Ian, include a brief explanation of what this is (plus citations)**]

Hydraulic traits were collected from SCBI and are summarized in Table 3. In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot. These were covered with opaque plastic bags, re-cut under water, and re-hydrated overnight before further analysis. Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using an R script (**details**). $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA$ was calculated as the percent loss of area between fresh and dry leaves. $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume. We used the rapid determination method of ([Bartlett et al., 2012]) to estimate the turgor loss point ($\pi_{tlp}$). Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ ([Bartlett et al., 2012]). We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration. Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].

```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/species_table.csv")

#the percentage symbol has a double backslash because otherwise this writes out the latex code when making the table

kable(table3, booktabs=TRUE, caption = "**Table 3. Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements. Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

(**add description of climate data used in Fig. 1, NEON vertical profiles**)

*Identification of drought years*

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where tree growth was substantially reduced and where peak growing season climatic conditions were among the driest of the time period. To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011]. Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years. Candidate drought years were defined if >25% of the cored trees experienced <30% growth in a year compared to the previous 5 years. Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris etal., 2014). The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

*Analysis*  (**this needs work**)

Linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (candfull). All variables with dAICc > 2 were used as candidates for each scenario's best model (figure ???? (tested traits best))

### Results

*Focal droughts and their characteristics*

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species. Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from [@helcoski_growing_2019]. Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenandoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.* 
**The droughts differed in timing/duration/etc. .. The year 1966 was preceded by two relatively dry years... 1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry. **

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% in 1977, and #% in 1999.

*Tree size and drought resistance*

Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4). When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5). We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/univariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5. Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  kable_styling()
```

\raggedright

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2**e**). Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1). When considered alone, $CP$ had a significant influence only in the 1977 drought, during which trees with dominant $CP$ had the lowest $R$. When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed trees had the lowest $R$. Crown position was included in **roughly half** of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees. Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.

In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).  [**biological temperature?- probably drop**]

![**Figure 2. Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), (c) air temperature ($T_{air}$), and (d) biological temperature ($T_{biological}$)averaged over each month of the peak growing season (May-August) from YEAR-YEAR. In these plots, heights are slightly offset for visualization purposes. Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$. In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals. In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}

>>>>

---
title: "Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
output:
  pdf_document:
    fig_caption: yes
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{dcolumn} #for sidewaystable
---

@@

### Supplementary Information
```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```
